MultifidelityTrajectoryOptimization

### 2025-02-17 artefacts/paper.tex
clean up write-up till control variate/monte carlo

\documentclass{article}
\usepackage[a4paper, total={7in, 8in}]{geometry}
\usepackage{graphicx} % Required for inserting images
\usepackage{amsmath} % Required for some math elements
\usepackage{hyperref} % Required for hyperlinks
\usepackage{amssymb} % Required for probability symbols

@@

\section{Problem Formulation}
This secition describes how the state space model of the dynamics is set up, along with the controller design and cost function for the trajectory optimization problem.
\subsection{Mass-Spring-Damper System}
For this paper, a simplified trajectory optimization problem is considered that is easy to understand and analyze.
The objective is to solve a reference tracking problem for the position and velocity of a simple mass-spring-damper system.
The mass-spring-damper system has the following parameters:
\begin{itemize}
  \item $m = 5$ g : Mass
  \item $k = 2$ N/m : Spring constant
  \item $c = 0.5$ Ns/m : Damping coefficient
    % \item Initial position: $x_0 = 0$ m
    % \item Initial velocity: $\dot{x}_0 = 0$ m/s
\end{itemize}

@@

\begin{equation}
  m\ddot{x} = F - c\dot{x} - kx
  \implies \ddot{x} = \frac{F}{m} - \frac{c}{m}\dot{x} - \frac{k}{m}x
\end{equation}

The position $x$ and velocity $\dot{x}$ of the mass-spring-damper system can be changed using the control input $u$,
which applies the force $F$ on the mass. This yields the following dynamics

\begin{equation}
  \begin{bmatrix}
    \dot{x} \\
    \ddot{x}
  \end{bmatrix} =
  \begin{bmatrix}
    0 & 1 \\
    -\frac{k}{m}  -\frac{c}{m}
  \end{bmatrix}
  \begin{bmatrix}
    x \\
    \dot{x}
  \end{bmatrix} +
  \begin{bmatrix}
    0 \\
    \frac{1}{m}
  \end{bmatrix} u
\end{equation}

For a closed-loop system, the output $y$ of the system needs to be observed, which is both the position and velocity in this case.

\begin{equation}
  y =
  \begin{bmatrix}
    1 & 0 \\
    0 & 1
  \end{bmatrix}
  \begin{bmatrix}
    x \\
    \dot{x}
  \end{bmatrix}
\end{equation}

This yields an LTI system in the form of $\dot{x} = Ax + Bu$ and $y = Cx$, where

\begin{equation}
  \begin{aligned}
    A &=
    \begin{bmatrix}
      0 & 1 \\
      -\frac{k}{m} & -\frac{c}{m}
    \end{bmatrix} \\
    B &=
    \begin{bmatrix}
      0 \\
      \frac{1}{m}
    \end{bmatrix} \\
    C &=
    \begin{bmatrix}
      1 & 0 \\
      0 & 1
    \end{bmatrix}
  \end{aligned}
\end{equation}

\subsection{LQR Controller} \label{lqr_ctrl}
We set up an LQR controller to track the reference position and velocity of the mass-spring-damper system.
An LQR controller was chosen because it has a quadratic cost function that can be minimized analytically and gives us a
good reference for the trajectory optimization problem.

To set up the LQR controller, we need to define the cost function of the system, which consists of:

\begin{itemize}
  \item $u_0, u_1 ... u_{N-1}$ : control inputs at each timestep of the trajectory
  \item $J$ : total cost to be minimized
  \item $N$ : time horizon
  \item $x_k$ : state vector at time step $k$
  \item $u_k$ : control input vector at time step $k$
  \item $\Delta u_k = u_k - u_{k-1}$ : change in control input at time step $k$
  \item $r_k$ : reference output at time step $k$
  \item $e_k = y_k - r_k = Cx - r_k$ : error in the output at time step $k$
  \item $Q_e$ : error cost matrix
  \item $R$ : control cost matrix
\end{itemize}
The LQR controller minimizes the cost function

\begin{equation}
  \min_{u_0, u_1 ... u_{N-1}}J = \sum_{k=0}^{N-1} (e_k^T Q_e e_k + \Delta u_k^T R \Delta u_k)
\end{equation}

\subsection{Stochastic LQR Controller} \label{stochastic_lqr}
We will consider the case where the initial state $x_0$ is stochastic, with mean $\mu$ and covariance $\Sigma$.
The control input vector $U$ is still deterministic.

In this case, the state vector $X$ is stochastic, owing to the the initial condition $x_0$ being a vector of random variables. The state propagation equations are now expectations

\begin{equation}
  \begin{aligned}
    \mathbb{E}[x_1] &= \mathbb{E}[A x_0 + B u_0] = A \mathbb{E}[x_0] + B u_0 \\
    \mathbb{E}[x_2] &= \mathbb{E}[A x_1 + B u_1] = A \mathbb{E}[x_1] + B u_1 = A^2 \mathbb{E}[x_0] + A B u_0 + B u_1 \\
    \vdots \\
    \mathbb{E}[x_N] &= \mathbb{E}[A x_{N-1} + B u_{N-1}] = A \mathbb{E}[x_{N-1}] + B u_{N-1} \\
    &= A^N \mathbb{E}[x_0] + A^{N-1} B u_0 + A^{N-2} B u_1 + \ldots + B u_{N-1} \\
  \end{aligned}
\end{equation}

The matrices $S$ and $M$ remain the same, and we have the following expression for the expected value of the state vector $X$

$$
\mathbb{E}[X] = S U + M \mathbb{E}[x_0]
$$

The cost matrices $\bar{Q}$ and $\bar{R}$ are also the same. The cost function $J$ is now a random variable,
with an expected value $\mathbb{E}[J]$ and variance $Var[J]$.

The expected value of the cost function over all possible realizations of the initial conditions is
$$
\begin{aligned}
  \mathbb{E}[J] &= \mathbb{E}[x_N^T P x_N + \sum_{k=0}^{N-1} (x_k^T Q x_k + u_k^T R u_k)] \\
  &= \mathbb{E}[X^T \bar{Q} X + U^T \bar{R} U + x_0^T Q x_0] \\
  &= \mathbb{E}[X^T \bar{Q} X] + \mathbb{E}[U^T \bar{R} U] + \mathbb{E}[x_0^T Q x_0] \\
  &= \mathbb{E}[(S U + M x_0)^T \bar{Q} (S U + M x_0)] + U^T \bar{R} U + \mathbb{E}[x_0^T Q x_0] \\
  &= \mathbb{E}[U^T S^T \bar{Q} S U + U^T S^T \bar{Q} M x_0 + x_0^T M^T \bar{Q} S U + x_0^T M^T \bar{Q} M x_0] + U^T \bar{R} U + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
  &= U^T S^T \bar{Q} S U + \mathbb{E}[U^T S^T \bar{Q} M x_0] + \mathbb{E}[x_0^T M^T \bar{Q} S U] + \mathbb{E}[x_0^T M^T \bar{Q} M x_0] + U^T \bar{R} U + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
  &= U^T (S^T \bar{Q} S + \bar{R}) U + U^T S^T \bar{Q} M \mathbb{E}[x_0] + \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T M^T \bar{Q} M x_0] + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
  &= U^T (S^T \bar{Q} S + \bar{R}) U + \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T] M^T \bar{Q} M \mathbb{E}[x_0] + \text{tr}(M^T \bar{Q} M \text{Cov}[x_0]) \hdots \\
  & \quad + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
  &= U^T (S^T \bar{Q} S + \bar{R}) U + 2 \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T] (M^T \bar{Q} M + Q) \mathbb{E}[x_0] + \text{tr}((M^T \bar{Q} M + Q) \text{Cov}[x_0]) \\
\end{aligned}
$$

Assuming $\mathbb{E}[x_0]$ and $\text{Cov}[x_0]$ is calculated using a Monte Carlo estimator, we can write the cost function as

@@

$$
\Delta J = 2 H U + 2 q^T
$$

@@

$$
\begin{aligned}
  \Delta J &= 2 H U + 2 q^T = 0 \\
  \implies U^* &= -H^{-1} q
\end{aligned}
$$

% ref for quadratic form: https://en.wikipedia.org/wiki/Quadratic_form_(statistics)

The cost function can be rewritten as

$$
\begin{aligned}
  J &= U^T(S^T \bar{Q} S + \bar{R}) U + 2 x_0^T M^T \bar{Q} S U + x_0^T(M^T \bar{Q} M + Q) x_0 \\
  \text{Let}& \\
  K &= U^T(S^T \bar{Q} S + \bar{R}) U \\
  L &= 2 M^T \bar{Q} S U \\
  N &= M^T \bar{Q} M + Q = N^T \\
  &\text{(quadratic multiplication by diagnonal($\bar{Q}$) results in a symmetric, $Q$ is symmetric)} \\
  \text{Then}& \\
  J &= K + x_0^T L + x_0^T N x_0 \\
\end{aligned}
$$

Then, the expectation and variance of the cost function is as follows. Take note that the random variable here is $x_0$ and $\mathbb{E}[x_0]$ is a scalar.

% ref:
% - https://www.math.uwaterloo.ca/~hwolkowi/matrixcookbook.pdf
% - https://math.stackexchange.com/questions/2163694/expectation-of-quartic-form-for-multivariate-gaussian/4247240#4247240
% - https://math.stackexchange.com/questions/1302882/variance-of-a-quadratic-form

\begin{equation}
  \begin{aligned}
    \mathbb{E}[J] &= \mathbb{E}[K] + \mathbb{E}[x_0^T L] + \mathbb{E}[x_0^T N x_0] \\
    &= K + \mathbb{E}[x_0^T] L + \mathbb{E}[x_0^T] N \mathbb{E}[x_0] + \text{tr}(N \text{Cov}[x_0]) \\
    Var[J] &= Var[K + x_0^T L + x_0^T N x_0] \\
    &\text{note: $K$ is deterministic, so its variance is 0} \\
    & \text{note: $x_0^T L$ is a scalar, $x_0^T L = L^T x_0$} \\
    &= Var[L^T x_0 + x_0^T N x_0] \\
    &= 2\text{tr}(N \text{Cov}[x_0] N \text{Cov}[x_0]) + 4 \mathbb{E}[x_0^T]N\text{Cov}[x_0]N\mathbb{E}[x_0] + 4 L^T \text{Cov}[x_0] N \mathbb{E}[x_0] + L^T \text{Cov}[x_0] L \\
  \end{aligned}
\end{equation}

@@

\begin{equation}
  \begin{aligned}
    \text{Cov}[x_0^T L_1, x_0^T L_2] &= \text{Cov}[L_1^T x_0, L_2^T x_0] \text{ transpose of a scalar is itself} \\
    &= L_1^T \text{Cov}[x_0] L_2 \\
  \end{aligned}
\end{equation}

\begin{equation}
  % - https://www.researchgate.net/publication/243770190_Quadratic_Forms_in_Random_Variables P74, 3.2d.9
  \begin{aligned}
    \text{Cov}[x_0^T L_1, x_0^T N_2 x_0] &= \text{Cov}[x_0^T N_2 x_0, L_1^T x_0] \text{ transpose of a scalar is itself} \\
    &= 2 \mathbb{E}[x_0^T] N_2 \text{Cov}[x_0] L_1 \\
  \end{aligned}
\end{equation}

\begin{equation}
  % - https://www.researchgate.net/publication/243770190_Quadratic_Forms_in_Random_Variables P74, 3.2d.9
  \begin{aligned}
    \text{Cov}[x_0^T N_1 x_0, x_0^T L_2] &= \text{Cov}[x_0^T N_1 x_0, L_2^T x_0] \text{ transpose of a scalar is itself} \\
    &= 2 \mathbb{E}[x_0^T] N_1 \text{Cov}[x_0] L_2 \\
  \end{aligned}
\end{equation}

\begin{equation}
  % - https://www.researchgate.net/publication/243770190_Quadratic_Forms_in_Random_Variables P75, 3.2d.12
  \begin{aligned}
    \text{Cov}[x_0^T N_1 x_0, x_0^T N_2 x_0] &= 2 \text{tr}(N_1 \text{Cov}[x_0] N_2 \text{Cov}[x_0]) + 4 \mathbb{E}[x_0^T]N_1\text{Cov}[x_0]N_2\mathbb{E}[x_0] \\
  \end{aligned}
\end{equation}

@@

\section{Monte Carlo with Control Variates}

@@

\begin{itemize}
  \item $S_n^{MC}(J_h(u_h, x_0))$: MC estimate of the high fidelity cost function
  \item $S_n^{CV}(J_h, J_l)$: CV estimate with $u_{hla}$ at the current iteration
  \item $S_n^{CV_{max}}(J_h, J_l)$:CV estimate with $u_{hla}$ at the iteration with the highest correlation
\end{itemize}

>>>>

\documentclass{article}
\usepackage[a4paper, total={7in, 8in}]{geometry}
\usepackage{graphicx} % Required for inserting images
\usepackage{amsmath} % Required for some math elements
\usepackage{hyperref} % Required for hyperlinks
\usepackage{amssymb} % Required for probability symbols
\usepackage{algorithm}
\usepackage{algpseudocode}

@@

\section{Problem Formulation}
This secition describes how the state space model of the dynamics is set up, along with the controller design and cost function for the trajectory optimization problem.
\subsection{Mass-Spring-Damper System}
For this paper, a simplified trajectory optimization problem is considered for ease of modelling.
The objective is to solve a reference tracking problem for the position and velocity of a simple mass-spring-damper system.
The mass-spring-damper system has the following parameters:

\begin{itemize}
  \item $m = 5$ g : Mass
  \item $k = 2$ N/m : Spring constant
  \item $c = 0.5$ Ns/m : Damping coefficient
    % \item Initial position: $x_0 = 0$ m
    % \item Initial velocity: $\dot{x}_0 = 0$ m/s
\end{itemize}

@@

$$
\begin{aligned}
  m\ddot{x} &= F - c\dot{x} - kx \\
  \implies \ddot{x} &= \frac{F}{m} - \frac{c}{m}\dot{x} - \frac{k}{m}x
\end{aligned}
$$

We have the state as $[x \dot{x}]^T$ containing the position and velocity of the mass, both of which are observed.
The control input $u=F$ is applied to the mass, yielding the following state space model:

$$
\begin{aligned}
  \begin{bmatrix}
    \dot{x} \\
    \ddot{x}
  \end{bmatrix} &=
  \begin{bmatrix}
    0 & 1 \\
    -\frac{k}{m}  -\frac{c}{m}
  \end{bmatrix}
  \begin{bmatrix}
    x \\
    \dot{x}
  \end{bmatrix} +
  \begin{bmatrix}
    0 \\
    \frac{1}{m}
  \end{bmatrix} u \\
  y &=
  \begin{bmatrix}
    1 & 0 \\
    0 & 1
  \end{bmatrix}
  \begin{bmatrix}
    x \\
    \dot{x}
  \end{bmatrix}
\end{aligned}
$$

Thus, we have a standard continuous-time LTI system $\dot{x} = A_cx + B_cu$ and $y = C_cx$, where

\begin{equation}
  \begin{aligned}
    A_c &=
    \begin{bmatrix}
      0 & 1 \\
      -\frac{k}{m} & -\frac{c}{m}
    \end{bmatrix} \\
    B_c &=
    \begin{bmatrix}
      0 \\
      \frac{1}{m}
    \end{bmatrix} \\
    C_c &=
    \begin{bmatrix}
      1 & 0 \\
      0 & 1
    \end{bmatrix}
  \end{aligned}
\end{equation}

\subsection{LQR Controller} \label{lqr_ctrl}
The continuous-time system can be converted to discrete time using such that $x_{k+1} = A x_k + B u_k$ and $y_k = C x_k$ for a chosen sampling time $T_s$.
The $A$ and $B$ matrices for the discrete time system are then used to formulate the cost function for the LQR problem.

We set up an LQR controller to track the reference position and velocity of the mass-spring-damper system.
An LQR controller was chosen because it has a quadratic cost function that can be minimized analytically and gives us a
good reference for the trajectory optimization problem.

The LQR controller minimizes the cost function

$$
\min_{u_0, u_1 ... u_{N-1}}J = \sum_{k=0}^{N-1} (e_k^T Q_e e_k + \Delta u_k^T R \Delta u_k)
$$

Where

\begin{itemize}
  \item $\Delta u_0, \Delta u_1 ... \Delta u_{N-1}$ : control inputs at each timestep of the trajectory
  \item $J$ : total cost to be minimized
  \item $N$ : time horizon
  \item $x_k$ : state vector at time step $k$
  \item $u_k$ : control input vector at time step $k$
  \item $\Delta u_k = u_k - u_{k-1}$ : change in control input at time step $k$
  \item $r_k$ : reference output at time step $k$
  \item $e_k = y_k - r_k = Cx - r_k$ : error in the output at time step $k$
  \item $Q_e$ : error cost matrix
  \item $R$ : control cost matrix
\end{itemize}

Note that we can add a terminal penalty using the DARE solution for the LQR problem, but for simplicity we will not consider it here.

To track the error $e_k$, we extend the state vector $x_k$ to include the control input $u_{k-1}$ and the reference output $r_k$.

$$
x_k^{ext} =
\begin{bmatrix}
  x_k \\
  u_{k-1} \\
  r_k
\end{bmatrix}
$$

With the resulting state space model

\begin{equation}
  \begin{aligned}
    x_{k+1}^{ext} &=
    \begin{bmatrix}
      A & B & 0 \\
      0 & \mathbb{I}_{n_u \times n_u} & 0 \\
      0 & 0 & \mathbb{I}_{n_r \times n_r}
    \end{bmatrix}
    \begin{bmatrix}
      x_k \\
      u_{k-1} \\
      r_k
    \end{bmatrix} +
    \begin{bmatrix}
      B \\
      \mathbb{I}_{n_u \times n_u} \\
      0
    \end{bmatrix} \Delta u_k \\
  \end{aligned}
\end{equation}

with

$$
\begin{aligned}
  A^{ext} &=
  \begin{bmatrix}
    A & B & 0 \\
    0 & \mathbb{I}_{n_u \times n_u} & 0 \\
    0 & 0 & \mathbb{I}_{n_r \times n_r}
  \end{bmatrix} \\
  B^{ext} &=
  \begin{bmatrix}
    B \\
    \mathbb{I}_{n_u \times n_u} \\
    0
  \end{bmatrix}
\end{aligned}
$$

Similary, $Qe$ with the extended state $x_k^{ext}$ transforms to $Q$ such that $e_k^T Q_e e_k = x_k^{ext^T} Q x_k^{ext}$.

$$
\begin{aligned}
  e_k^T Q_e e_k &= (y_k - r_k)^T Q_e (y_k - r_k) \\
  &= \left(
    \begin{bmatrix} C & 0 & -\mathbb{I}_{n_r \times n_r}
    \end{bmatrix}
    \begin{bmatrix}
      x_k \\
      u_{k-1} \\
      r_k
  \end{bmatrix}\right)^T Q_e
  \left(
    \begin{bmatrix} C & 0 & -\mathbb{I}_{n_r \times n_r}
    \end{bmatrix}
    \begin{bmatrix}
      x_k \\
      u_{k-1} \\
      r_k
  \end{bmatrix}\right) \\
  &= x_k^{ext^T}
  \begin{bmatrix} C & 0 & -\mathbb{I}_{n_r \times n_r}
  \end{bmatrix}^T Q_e
  \begin{bmatrix} C & 0 & -\mathbb{I}_{n_r \times n_r}
  \end{bmatrix} x_k^{ext} \\
\end{aligned}
$$

\begin{equation}
  Q =
  \begin{bmatrix} C & 0 & -\mathbb{I}_{n_r \times n_r}
  \end{bmatrix}^T Q_e
  \begin{bmatrix} C & 0 & -\mathbb{I}_{n_r \times n_r}
  \end{bmatrix}
\end{equation}

For simplicity, from now on we refer to $x_k^{ext}$ as $x_k$, $A^{ext}$ as $A$ and $B^{ext}$ as $B$.
The cost function can now be written as

$$
J = \sum_{k=0}^{N-1} x_k^T Q x_k + \Delta u_k^T R \Delta u_k
$$

Using the state transition formula

$$
\begin{aligned}
  x_1 &= A x_0 + B u_0 \\
  x_2 &= A x_1 + B u_1 = A(A x_0 + B u_0) + B u_1 = A^2 x_0 + A B u_0 + B u_1 \\
  \vdots \\
  x_N &= A x_{N-1} + B u_{N-1} = A^{N-1} x_0 + A^{N-2} B u_0 + A^{N-3} B u_1 + \ldots + B u_{N-1} \\
  \implies x_k &= A^k x_0 + \sum_{i=0}^{k-1} A^{k-1-i} B u_i
\end{aligned}
$$

We construct the matrices $X$, $U$, $S$ and $M$ such that $X = SU + Mx_0$.

$$
\begin{aligned}
  X &=
  \begin{bmatrix}
    x_1^{ext} \\
    x_2^{ext} \\
    \vdots \\
    x_N^{ext}
  \end{bmatrix} & \in \mathbb{R}^{Nn_x \times 1} \\
  U &=
  \begin{bmatrix}
    \Delta u_0 \\
    \Delta u_1 \\
    \vdots \\
    \Delta u_{N-1}
  \end{bmatrix} & \in \mathbb{R}^{Nn_u \times 1} \\
  S &=
  \begin{bmatrix}
    B & 0 & \ldots & 0 \\
    AB & B & \ldots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    A^{N-1} B & A^{N-2} B & \ldots & B
  \end{bmatrix} & \in \mathbb{R}^{Nn_x \times Nn_u} \\
  M &=
  \begin{bmatrix}
    A \\
    A^2 \\
    \vdots \\
    A^N
  \end{bmatrix} & \in \mathbb{R}^{Nn_x \times n_x}
\end{aligned}
$$

Similarly, the matrices $\bar{Q}$ and $\bar{R}$ are constructed

$$
\begin{aligned}
  \bar{Q} &=
  \begin{bmatrix}
    Q & 0 & \ldots & 0 \\
    0 & Q & \ldots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & \ldots & 0
  \end{bmatrix} \text{with size } Nn_x \times Nn_x \\
  \bar{R} &=
  \begin{bmatrix}
    R & 0 & \ldots & 0 \\
    0 & R & \ldots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & \ldots & R
  \end{bmatrix} \text{with size } Nn_u \times Nn_u
\end{aligned}
$$

The cost function can then be written as a quadratic function of the control inputs $U$

\begin{equation}
  \begin{aligned}
    J &= \sum_{k=0}^{N-1} x_k^T Q x_k + \Delta u_k^T R \Delta u_k \\
    % \text{ (extra $x_0$ term because it is not in $X$)} \\
    &= X^T \bar{Q} X + U^T \bar{R} U + x_0^T Q x_0 \\
    &= (S U + M x_0)^T \bar{Q} (S U + M x_0) + U^T \bar{R} U + x_0^T Q x_0 \\
    % & \text{note } (A + B)^T = A^T + B^T \text{ and } (AB)^T = B^T A^T \\
    &= (U^T S^T + x_0^T M^T) \bar{Q} (S U + M x_0) + U^T \bar{R} U + x_0^T Q x_0 \\
    % & \text{note } A(B + C) = AB + AC \\
    &= (U^T S^T + x_0^T M^T) (\bar{Q} S U + \bar{Q} M x_0) + U^T \bar{R} U + x_0^T Q x_0 \\
    &= U^T S^T \bar{Q} S U + U^T S^T \bar{Q} M x_0 + x_0^T M^T \bar{Q} S U + x_0^T M^T \bar{Q} M x_0 + U^T \bar{R} U + x_0^T Q x_0 \\
    % & \text{note } \bar{Q} \text{ is symmetric so it doesnt need to be transposed} \\
    &= U^T S^T \bar{Q} S U + (x_0^T M^T \bar{Q} S U)^T + x_0^T M^T \bar{Q} S U + x_0^T M^T \bar{Q} M x_0 + U^T \bar{R} U + x_0^T Q x_0 \\
    % & \text{note } x_0^T M^T \bar{Q} S U \text{ is a scalar(check its dimensions), so its transpose is itself} \\
    &= U^T S^T \bar{Q} S U + x_0^T M^T \bar{Q} S U + x_0^T M^T \bar{Q} S U + x_0^T M^T \bar{Q} M x_0 + U^T \bar{R} U + x_0^T Q x_0 \\
    % & \text{combining all quadratic and repeated terms} \\
    &= U^T (S^T \bar{Q} S + \bar{R}) U + 2 x_0^T M^T \bar{Q} S U + x_0^T(M^T \bar{Q} M + Q) x_0\\
  \end{aligned}
\end{equation}

Let

$$
\begin{aligned}
  H &= S^T \bar{Q} S + \bar{R} = H^T \\
  &\text{(quadratic multiplication by diagnonal($\bar{Q}$) results in a symmetric, R is symmetric)} \\
  q &= (x_0^T M^T \bar{Q} S)^T = S^T \bar{Q} M x_0 \\
  c &= x_0^T (M^T \bar{Q} M + Q) x_0
\end{aligned}
$$

@@

$$
\frac{\partial J}{\partial U} = 2 H U + 2 q^T
$$

@@

\begin{equation}
  \begin{aligned}
    \frac{\partial J}{\partial U} &= 2 H U + 2 q^T = 0 \\
    \implies U^* &= -H^{-1} q
  \end{aligned}
\end{equation}

\subsection{Stochastic LQR Controller} \label{stochastic_lqr}

We consider the case where the initial state $x_0$ is stochastic, with mean $\mu$ and covariance $\Sigma$.
Hence, the state vector $X$ is stochastic, since it is propagating the randomness of $x_0$ forward.
The control input vector $U$ is deterministic. The cost function $J$ is now a random variable,
with an expected value $\mathbb{E}[J]$ and variance $Var[J]$.

\subsubsection{Solution of the Stochastic LQR controller}

The expected value of the cost function over all possible realizations of $x_0$ is

\begin{equation}
  \begin{aligned}
    \mathbb{E}[J] &= \mathbb{E}[\sum_{k=0}^{N-1} x_k^T Q x_k + \Delta u_k^T R \Delta u_k] \\
    &= \mathbb{E}[X^T \bar{Q} X + U^T \bar{R} U + x_0^T Q x_0] \\
    &= \mathbb{E}[X^T \bar{Q} X] + \mathbb{E}[U^T \bar{R} U] + \mathbb{E}[x_0^T Q x_0] \\
    &= \mathbb{E}[(S U + M x_0)^T \bar{Q} (S U + M x_0)] + U^T \bar{R} U + \mathbb{E}[x_0^T Q x_0] \\
    &= \mathbb{E}[U^T S^T \bar{Q} S U + U^T S^T \bar{Q} M x_0 + x_0^T M^T \bar{Q} S U + x_0^T M^T \bar{Q} M x_0] + U^T \bar{R} U + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
    &= U^T S^T \bar{Q} S U + \mathbb{E}[U^T S^T \bar{Q} M x_0] + \mathbb{E}[x_0^T M^T \bar{Q} S U] + \mathbb{E}[x_0^T M^T \bar{Q} M x_0] + U^T \bar{R} U + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
    &= U^T (S^T \bar{Q} S + \bar{R}) U + U^T S^T \bar{Q} M \mathbb{E}[x_0] + \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T M^T \bar{Q} M x_0] + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
    &= U^T (S^T \bar{Q} S + \bar{R}) U + \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T] M^T \bar{Q} M \mathbb{E}[x_0] + \text{tr}(M^T \bar{Q} M \text{Cov}[x_0]) \dots \\
    & \quad + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
    &= U^T (S^T \bar{Q} S + \bar{R}) U + 2 \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T] (M^T \bar{Q} M + Q) \mathbb{E}[x_0] + \text{tr}((M^T \bar{Q} M + Q) \text{Cov}[x_0]) \\
  \end{aligned}
\end{equation}

Let

The expectation of cost function can be written as

$$
\mathbb{E}[J] = U^T H U + 2 q^T U + c
$$

$$
\frac{\partial \mathbb{E}[J]}{\partial U} = 2 H U + 2 q^T
$$

Assuming minimum exists, we can set the gradient to zero to get the optimal control input $U^*$ that minimizes the expected value of the cost function

\begin{equation}
  \begin{aligned}
    \frac{\partial \mathbb{E}[J]}{\partial U} &= 2 H U + 2 q^T = 0 \\
    \implies U^* &= -H^{-1} q
  \end{aligned}
\end{equation}

\subsubsection{Variance and Covariance of cost function in terms of initial state}

% ref for quadratic form: https://en.wikipedia.org/wiki/Quadratic_form_(statistics)
% ref:
% - https://www.math.uwaterloo.ca/~hwolkowi/matrixcookbook.pdf
% - https://math.stackexchange.com/questions/2163694/expectation-of-quartic-form-for-multivariate-gaussian/4247240#4247240
% - https://math.stackexchange.com/questions/1302882/variance-of-a-quadratic-form

The expected value of the cost function can be rewritten as

$$
\mathbb{E}[J] = K + \mathbb{E}[x_0^T] L + \mathbb{E}[x_0^T] N \mathbb{E}[x_0] + \text{tr}(N \text{Cov}[x_0]) \\
$$

Where

$$
\begin{aligned}
  x_0 &\sim \mathcal{N}(\mu, \Sigma) \\
  K &= U^T(S^T \bar{Q} S + \bar{R}) U \\
  L &= 2 M^T \bar{Q} S U \\
  N &= M^T \bar{Q} M + Q = N^T \\
  &\text{(quadratic multiplication by diagnonal($\bar{Q}$) results in a symmetric, $Q$ is symmetric)} \\
\end{aligned}
$$

Using this, the variance can be calculated.

\begin{equation}
  \begin{aligned}
    Var[J] &= Var[K + x_0^T L + x_0^T N x_0] \\
    &\text{note: $K$ is deterministic, so its variance is 0} \\
    & \text{note: $x_0^T L$ is a scalar, $x_0^T L = L^T x_0$} \\
    &= Var[L^T x_0 + x_0^T N x_0] \\
    &= 2\text{tr}(N \text{Cov}[x_0] N \text{Cov}[x_0]) + 4 \mathbb{E}[x_0^T]N\text{Cov}[x_0]N\mathbb{E}[x_0] + 4 L^T \text{Cov}[x_0] N \mathbb{E}[x_0] + L^T \text{Cov}[x_0] L \\
  \end{aligned}
\end{equation}

@@

$$
\begin{aligned}
  \text{Cov}[x_0^T L_1, x_0^T L_2] &= \text{Cov}[L_1^T x_0, L_2^T x_0] \text{ transpose of a scalar is itself} \\
  &= L_1^T \text{Cov}[x_0] L_2 \\
\end{aligned}
$$

$$
% - https://www.researchgate.net/publication/243770190_Quadratic_Forms_in_Random_Variables P74, 3.2d.9
\begin{aligned}
  \text{Cov}[x_0^T L_1, x_0^T N_2 x_0] &= \text{Cov}[x_0^T N_2 x_0, L_1^T x_0] \text{ transpose of a scalar is itself} \\
  &= 2 \mathbb{E}[x_0^T] N_2 \text{Cov}[x_0] L_1 \\
\end{aligned}
$$

$$
% - https://www.researchgate.net/publication/243770190_Quadratic_Forms_in_Random_Variables P74, 3.2d.9
\begin{aligned}
  \text{Cov}[x_0^T N_1 x_0, x_0^T L_2] &= \text{Cov}[x_0^T N_1 x_0, L_2^T x_0] \text{ transpose of a scalar is itself} \\
  &= 2 \mathbb{E}[x_0^T] N_1 \text{Cov}[x_0] L_2 \\
\end{aligned}
$$

$$
% - https://www.researchgate.net/publication/243770190_Quadratic_Forms_in_Random_Variables P75, 3.2d.12
\begin{aligned}
  \text{Cov}[x_0^T N_1 x_0, x_0^T N_2 x_0] &= 2 \text{tr}(N_1 \text{Cov}[x_0] N_2 \text{Cov}[x_0]) + 4 \mathbb{E}[x_0^T]N_1\text{Cov}[x_0]N_2\mathbb{E}[x_0] \\
\end{aligned}
$$

@@

\section{Monte Carlo with Control Variates}
TODO: explain why we need control variates

@@

\begin{itemize}
  \item $S_n^{MC}(J_h(U_h, x_0))$: MC estimate of the high fidelity cost function
  \item $S_n^{CV}(J_h(U_h, x_0), J_l(U_{hla}, x_0))$: CV estimate with $u_{hla}$ at the current iteration
  \item $S_n^{CV_{max}}(J_h(U_h, x_0), J_l(U_{hla}^{max}, x_0))$:CV estimate where $u_{hla}^{max}$ is at the iteration with the highest correlation.
\end{itemize}

This is how the whole routine works, step-by-step:

Initialization:
\begin{itemize}
  \item $U_l^* \gets$ Analytical solution of low fidelity model
  \item $U_{h_0} \gets T_{lh} U_l^*$
\end{itemize}

% For $S_n^{CV}$ the numerical optimizer is run for $i^{max}$ iterations, where in each iteration $i$, the objective is calculated:
% \begin{equation}
%   S_n^{CV}(J_h, J_l) = S_n^{MC}(J_h(U, x_0)) + \alpha (S_n^{MC}(J_l(U_{hla}, x_0)) - \mathbb{E}[J_l(U_{hla}, x_0)])
% \end{equation}

% Where
% \begin{itemize}
%   \item $n$: Number of samples of $x_0$ used in the MC estimate to equal the cost of $S^{MC}(J_h(U, x_0))$ with all samples
%   \item $U_{hla} = T_{hla} U$
%   \item $\alpha = -\frac{\text{Cov}(J_h(U, x_0), J_l(U_{hla}, x_0))}{\text{Var}[J_l(U_{hla}, x_0)]}$ is calculated analytically
%   \item $\mathbb{E}[J_l(U_{hla}, x_0)]$ is calculated analytically
% \end{itemize}

% For $S_n^{CV_{max}}$ the numerical optimizer is run for $i^{max}$ iterations, where in each iteration $i$, the objective is calculated:
% \begin{equation}
%   S_n^{CV_{max}}(J_h, J_l) = S_n^{MC}(J_h(U, x_0)) + \alpha (S_n^{MC}(J_l(U_{hla}^{max}, x_0)) - \mathbb{E}[J_l(U_{hla}^{max}, x_0)])
% \end{equation}
